Universals of word order result from optimization of grammars for efficient communication

### 2019-03-17 second-draft/3-paper-names.tex
revising

In this work, we first provide a formalization of efficiency in the domain of word order, drawing on information-theoretic notions.
We then show that the word order of languages has evolved to optimize efficiency, and that efficiency optimization accounts both for the Greenbergian word order correlations and for Dependency Length Minimization.

@@

We model the process where a speaker communicates a message to a listener (Figure~\ref{fig:comm}) in Shannon's framework of information theory.
In this model, a transmitter encodes a message into a signal. The receiver decodes the signal, attempting to reconstruct the original message.
Applying this model to word order, we take the message to consist of entities, events, and a set of syntactic and semantic relations that hold between them.
Following a long tradition in formal and computational liguistics, we formalize these in the format of Dependency Grammar:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.
In Figure~\ref{fig:comm}, the message consists of a dependency structure with the words `saw', `Mary', `John', where Mary is the subject of the event denoted by `saw', and John is the object.

When uttering, the speaker needs to choose an ordering in which to order the words in this tree to generate a string of words.
Should the speaker do it as
Mary saw John (like in English)
or as 
Mary John saw (like in Japanese)

Depending on the language, the listener decodes this.

The listener receives this string of words.
By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
The dependency structure (including the labels on the arcs) specifies how the meanings of words are combined.
Therefore, a listener needs to recover the information provided in the dependency structure in order to understand a sentence correctly.
Consistent with Shannon's model, we assume a Bayesian listener who decodes dependency structures probabilistically~(Figure~\ref{fig:comm}).
If communication is successful, the listener can identify the intended structure with high confidence~(Figure~\ref{fig:comm}).

@@

@@

In our model, these will depend on the grammar:
Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance.
This reflects the \key{parseability} of the language.

The degree to which listeners can reconstruct dependency structures from an utterance is formalized as the amount of information that utterances $u$ provide about their underlying dependency structures $t$:
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.
This quantity formalizes Zipf's principle of listener effort and his Force of Diversification:
It can be maximized when trees can be decoded fully unambiguously from utterances.
Maximizing the information provided by utterances about meanings is a standard formalization of Zipf's principle of listener economy~\cite{frank2012predicting,regier2015word}.

@@

The \key{efficiency} of a language is a weighted combination
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Pars}} + \lambda R_\textit{Pred}
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification and for robustness to other choices).

@@

We first demonstrate that real languages are relatively efficient compared to random baselines. For each language, we generated ten baseline word order grammars for the language.
These grammars have systematic word order regularities similar to natural language, but do not exhibit any correlations among the orderings of different syntactic relations (compare Figure~\ref{fig:grammars}).
The majority of real languages in Figure~\ref{fig:pareto-plane} are below and to the left of their baseline equivalents, demonstrating that they are relatively high in predictability and/or parseability.
X languages improve over the baselines in parsability, Y improve in predictability.
Z improve in overall efficiency.
Pareto frontier

Figure~\ref{fig:pareto-plane} also shows the average position of optimized languages.
Languages appear to be attracted toward these points and away from the region of the baseline languages. We also see that several languages actually end up \emph{more} efficient than the computationally optimized languages.

\begin{figure}
    \centering
    \includegraphics[scale=.45]{../results/plane/pareto-plane-iso-best-balanced-legend.pdf}
    \caption{Predictability and parseability of 51 languages (red), indicated by ISO codes, compared to baseline word order grammars per language (blue distribution). Predictability and parseability scores are $z$-scored within language. Each point for a real language has a line pointing in the direction of the center of mass of its baselines. Unlabeled dots represent the centroid for real languages (red), baseline languages (blue), and languages optimized for predictability (purple), parseability (green), efficiency (orange), and dependency length (yellow). When a language is to the bottom--left of its baselines, this indicates that it is relatively optimal for efficiency.} % TODO colors, fontsize, etc could conceivably be made prettier
    \label{fig:pareto-plane}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale=.45]{../results/plane/pareto-plane-iso-best-balanced-legend.pdf}
    \caption{Predictability and parseability of 51 UD languages (red), indicated by ISO codes, compared to ten baseline word order grammars per language (green). Predictability and parseability scores are $z$-scored within language. Each point for a real language has a line pointing in the direction of the center of mass of its baselines. The green contour shows the density of baseline languages. Unlabeled dots represent the centroid for real languages (red), baseline languages (green), and languages optimized for predictability (yellow), parseability (pink), efficiency (blue), and dependency length (red). When a language is to the bottom--left of its baselines, this indicates that it is relatively optimal for efficiency.} % TODO colors, fontsize, etc could conceivably be made prettier
    \label{fig:pareto-plane}
\end{figure}

>>>>

The contribution of our work is to show  that the word order of languages has evolved to optimize efficiency, and that efficiency optimization accounts for the Greenbergian word order correlations.
%In this work, we first provide a formalization of efficiency in the domain of word order. %, drawing on information-theoretic notions.

@@

%\subsection{Counterfactual Grammars}
We model the process where a speaker communicates a message to a listener (Figure~\ref{fig:comm}) in Shannon's framework of information theory.
In this model, a transmitter encodes a message into a signal. The receiver decodes the signal, attempting to reconstruct the original message.
Applying this model to word order, we take the message to consist of entities, events, and a set of syntactic and semantic relations that hold between them.
Following a long tradition in formal and computational liguistics, we formalize these in the format of Dependency Grammar:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.
In Figure~\ref{fig:comm}, the message consists of a dependency structure with the words `saw', `Mary', `John', where Mary is the subject of the event denoted by `saw', and John is the object.

When uttering, the speaker needs to choose an ordering in which to order the words in this tree to generate a string of words.
Imagine the speaker wants to communicate that Mary saw John.
Should the speaker say (1) as in English, or (2) as in Japanese?

(1)
\begin{dependency}[theme=simple]
    \begin{deptext}
    Mary \& saw \& John \\
    \end{deptext}
    \depedge{2}{1}{nsubj}
    \depedge{2}{3}{obj}
\end{dependency}
(2)
\begin{dependency}[theme=simple]
    \begin{deptext}
    Mary \& John  \& saw \\
    \end{deptext}
    \depedge{3}{1}{nsubj}
    \depedge{3}{2}{obj}
\end{dependency}

The listener receives a sentence -- a linear string of words -- and decodes the underlying message.
Consistent with Shannon's model, we assume a Bayesian listener who decodes dependency structures probabilistically~(Figure~\ref{fig:comm}).
If communication is successful, the listener can identify the intended structure with high confidence~(Figure~\ref{fig:comm}).

%The listener receives this string of words.
%By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
%The dependency structure (including the labels on the arcs) specifies how the meanings of words are combined.
%Therefore, a listener needs to recover the information provided in the dependency structure in order to understand a sentence correctly.

@@

%\subsection{Efficiency in Word Order}
Zipf

@@

Efficiency is formalized as a weighted combination of terms representing the amount of information conveyed and the cost of communication.
%The \key{efficiency} of a language is a weighted combination
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Pars}} + \lambda R_\textit{Pred}
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. %In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification and for robustness to other choices).

In our model, these will depend on the grammar:
Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance.
This reflects the \key{parseability} of the language.
Maximizing the information provided by utterances about meanings is a standard formalization of Zipf's principle of listener economy~\cite{frank2012predicting,regier2015word}.
The degree to which listeners can reconstruct dependency structures from an utterance is formalized as the amount of information that utterances $u$ provide about their underlying dependency structures $t$:
\begin{equation}
	R_{Pars} := I[\mathcal{U},\mathcal{T}] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.
This quantity formalizes Zipf's Force of Diversification:
It can be maximized when trees can be decoded fully unambiguously from utterances.

@@

@@

We first demonstrate that real languages are relatively efficient compared to random baselines. For each language, we generated ten baseline word order grammars for the language.
These grammars have systematic word order regularities similar to natural language, but do not exhibit any correlations among the orderings of different syntactic relations (compare Figure~\ref{fig:grammars}).
In Figure~\ref{fig:pareto-plane}, we plot predictability and parsability of the grammars of 51 languages, together with the distribution of random baseline grammars, and the Pareto frontier defined by computationally optimized grammars.
The majority of real languages are above and to the right of their baseline equivalents, demonstrating that they are relatively high in predictability and/or parseability.
X languages improve over the baselines in parsability ($p < ...$ each), Y improve in predictability ($p < ...$ each).
Z improve in overall efficiency ($p < ...$ each).
Languages are attracted towards the Pareto frontier and away from the region of the baseline languages.
A few languages end up \emph{more} efficient than the computationally optimized languages.

%Figure~\ref{fig:pareto-plane} also shows the average position of optimized languages.
%Languages appear to be attracted toward these points and away from the region of the baseline languages. We also see that several languages actually end up \emph{more} efficient than the computationally optimized languages.

\begin{figure}
    \centering
    \includegraphics[scale=.45]{../results/plane/pareto-plane-iso-best-balanced-legend-viz.pdf}
    \caption{Predictability and parseability of 51 languages (red), indicated by ISO codes, compared to baseline word order grammars (blue distribution). Predictability and parseability scores are $z$-scored within language.} % TODO colors, fontsize, etc could conceivably be made prettier
    \label{fig:pareto-plane}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale=.45]{../results/plane/pareto-plane-iso-best-balanced-legend-viz.pdf}
    \caption{Predictability and parseability of 51 UD languages (red), indicated by ISO codes, compared to ten baseline word order grammars per language (green). Predictability and parseability scores are $z$-scored within language. Each point for a real language has a line pointing in the direction of the center of mass of its baselines. The green contour shows the density of baseline languages. Unlabeled dots represent the centroid for real languages (red), baseline languages (green), and languages optimized for predictability (yellow), parseability (pink), efficiency (blue), and dependency length (red). When a language is to the bottom--left of its baselines, this indicates that it is relatively optimal for efficiency.} % TODO colors, fontsize, etc could conceivably be made prettier
    \label{fig:pareto-plane}
\end{figure}
